\section{AI Protocol and Capability Definition}

\subsection{Protocol Role}

The AI Protocol defines how capabilities are described. It specifies:
\begin{itemize}
  \item \textbf{Providers:} entities offering AI or non-AI capabilities.
  \item \textbf{Models:} specific AI models with defined capabilities.
  \item \textbf{Capability Types:} categories of capabilities such as chat, embedding, and image generation.
  \item \textbf{Endpoints:} where capabilities can be invoked.
  \item \textbf{Parameters:} input and output schemas.
\end{itemize}

The protocol serves as the contract boundary between orchestration and execution: orchestration decides policy, while execution enforces invocation semantics.

The protocol is typically represented as structured manifests:
\begin{lstlisting}[language={},caption={Provider-centric protocol manifest organization.},label={lst:provider-manifest}]
provider/
|- metadata.yaml
\- models/
   |- model_a.yaml
   \- model_b.yaml
      \- capabilities/
         |- chat.yaml
         \- embedding.yaml
\end{lstlisting}

\subsection{Capability Definition}

Each capability is defined with:
\begin{itemize}
  \item \textbf{Name:} a semantic identifier such as \texttt{generate\_text} or \texttt{search\_web}.
  \item \textbf{Input Schema:} a structured input specification.
  \item \textbf{Output Schema:} a structured output specification.
  \item \textbf{Execution Semantics:} the invocation and runtime behavior contract.
  \item \textbf{Quality Metadata ($Q$):} measured quality fields emitted by execution.
  \item \textbf{Pricing Metadata ($\Pi$):} metering and billing fields emitted by execution.
  \item \textbf{Provider Metadata:} information about the provider offering this capability.
\end{itemize}

\begin{lstlisting}[language={},caption={Example capability definition.},label={lst:capability-definition}]
name: generate_text
description: Generate text based on a prompt
input_schema:
  type: object
  properties:
    prompt:
      type: string
      description: The input prompt
    max_tokens:
      type: integer
      description: Maximum tokens to generate
  required: [prompt]
output_schema:
  type: object
  properties:
    text:
      type: string
      description: Generated text
execution:
  endpoint: /v1/completions
  method: POST
quality:
  telemetry_fields: [execution_latency_ms, retry_count]
pricing:
  billing_model: per_token
  currency: USD
provider:
  name: Example Provider
  version: 1.0.0
\end{lstlisting}

\subsection{Provider-Centric Organization}

An important design decision is that capability descriptions remain provider-centric rather than capability-centric. The manifest structure follows:
\begin{center}
\texttt{Provider} $\rightarrow$ \texttt{Models} $\rightarrow$ \texttt{Capabilities}
\end{center}
rather than
\begin{center}
\texttt{Capability} $\rightarrow$ \texttt{Providers}.
\end{center}

The runtime can dynamically construct a capability index:
\begin{lstlisting}[language={},caption={Runtime-generated capability index view.},label={lst:capability-index}]
chat/
|- provider A
|- provider B
\- provider C
embedding/
|- provider A
\- provider D
\end{lstlisting}

This index can be rebuilt periodically and optimized based on user preferences, latency measurements, cost calculations, and reliability metrics.
